---
title: ""
---
 
# **Edge Rusher Analysis**
***  

### **Introduction and Variables** 
The following analysis seeks to comparatively analyze the top 20 edge rushers available in the 2021 NFL draft. Both qualitative and quantitative variables were used to measure player value. The table below summarizes these metrics:  
  
* **PFF Overall**: Overall PFF Grade
* **PFF Pass**: Pass Defense PFF Grade
* **PFF Run**: Run Defense PFF Grade
* **Stop Rate**: Stops per Snap Played
* **Tackle Rate**: Tackles per Snap Played
* **Assist Rate**: Assists per Snap Played
* **Forced Fumble Rate**: Forced Fumbles per Snap Played
* **Hit Rate**: QB Hits per Pass Play
* **Hurry Rate**: QB Hurries per Pass Play
* **Sack Rate**: Sacks per Pass Play

### **Method**  
PFF grades range from 60-100 and are presented largely unchanged throughout this analysis. The rate metrics were combined and blended to create two new metrics: Composite Pass and Composite Run. These two scores range from 0-100 and measure the overall production of each player relative to the entire player pool. Below are the components of each score:  
  
  * **Composite Pass**: Hit Rate, Hurry Rate, and Sack Rate
  * **Composite Run**: Stop Rate, Tackle Rate, Assist Rate, and Forced Fumble Rate
  
  ***
### **Results**  
  
**Player Values by Category**  
The table below includes the PFF grades and composite score for the entirety of the player pool. The table can be sorted by column simply by clicking on the column title, and the search bar can be used to find specific players.  
```{r, echo = FALSE, results='hide', message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)

# Get a list of draft eligible players
player_set <- read_csv('data/draft_eligible_top20.csv')
in_draft <- player_set$player

# Read in the stats
metrics_2016 <- read_csv('data/defense_summary_2016.csv')
metrics_2017 <- read_csv('data/defense_summary_2017.csv')
metrics_2018 <- read_csv('data/defense_summary_2018.csv')
metrics_2019 <- read_csv('data/defense_summary_2019.csv')
metrics_2020 <- read_csv('data/defense_summary_2020.csv')

# Add year
metrics_2016$year <- '2016'
metrics_2017$year <- '2017'
metrics_2018$year <- '2018'
metrics_2019$year <- '2019'
metrics_2020$year <- '2020'

# Set the columns to keep
clean_up_columns = c('player',
                     'player_id',
                     'position',
                     'team_name',
                     'year',               
                     'player_game_count',
                     'assists',
                     'batted_passes',
                     'forced_fumbles',
                     'grades_defense',
                     'grades_pass_rush_defense',
                     'grades_run_defense',
                     'grades_tackle',
                     'hits',
                     'hurries',
                     'interceptions',
                     'missed_tackles',
                     'pass_break_ups',
                     'sacks',
                     'snap_counts_defense',
                     'snap_counts_pass_rush',
                     'snap_counts_run_defense',
                     'stops',
                     'tackles',
                     'total_pressures')

# Reduce the width of the data frames
metrics_2016 <- metrics_2016[clean_up_columns]
metrics_2017 <- metrics_2017[clean_up_columns]
metrics_2018 <- metrics_2018[clean_up_columns]
metrics_2019 <- metrics_2019[clean_up_columns]
metrics_2020 <- metrics_2020[clean_up_columns]

# Combine the data
full_data <- rbind(metrics_2016,
      metrics_2017,
      metrics_2018,
      metrics_2019,
      metrics_2020)

# Filter the data
start_year <- 2016
end_year <- 2020
full_data <- full_data[ which(full_data$year >= start_year
                         & full_data$year <= end_year), ]

full_data[c(3)][which(full_data$player == "Jordan Smith"), ] <- "UAB"
full_data[c(2)][which(full_data$player == "Jordan Smith"), ] <- 43427

# Aggregate the data
agg_data <- full_data %>%
  group_by(player, player_id) %>%
  summarise(games = sum(player_game_count), 
            snaps_pass = sum(snap_counts_pass_rush),
            snaps_run = sum(snap_counts_run_defense),
            stops = sum(stops),
            tackles = sum(tackles),
            assists = sum(assists),
            batted_passes = sum(batted_passes),
            forced_fumbles = sum(forced_fumbles),
            grade_tot = mean(grades_defense),
            grade_pr = mean(grades_pass_rush_defense),
            grade_rd = mean(grades_run_defense),
            hits = sum(hits),
            hurries = sum(hurries),
            sacks = sum(sacks))

# Filter to minimum games
min_games <- 0
agg_data <- agg_data[ which(agg_data$games >= min_games), ]

# Add a total snaps column
agg_data$snaps <- agg_data$snaps_pass + agg_data$snaps_run

# Make a new df for rates
rate_data <- agg_data

# Adjust the columns and make them rate based
rate_data$stop_rate = rate_data$stops / rate_data$snaps
rate_data$tackle_rate = rate_data$tackles / rate_data$snaps
rate_data$assist_rate = rate_data$assists / rate_data$snaps
rate_data$batted_passes = rate_data$batted_passes / rate_data$snaps_pass
rate_data$forced_fumbles = rate_data$forced_fumbles / rate_data$snaps
rate_data$hit_rate = rate_data$hits / rate_data$snaps_pass
rate_data$hurry_rate = rate_data$hurries / rate_data$snaps_pass
rate_data$sack_rate = rate_data$sacks / rate_data$snaps_pass

# Fix errors
rate_data[is.na(rate_data)] <- 0

# Isolate just the columns we want
cols_to_drop <- c('stops',
                  'tackles',
                  'assists',
                  'batted_passes',
                  'forced_fumbles',
                  'hits',
                  'hurries',
                  'sacks')

rate_data <- rate_data[ , !(names(rate_data) %in% cols_to_drop)]

# Limit to only the desired player set
rate_data <- subset(rate_data, player %in% in_draft)

# MinMax Scale
# (x-min)/(max-min)
scaled_data <- rate_data

stop_max <- max(scaled_data$stop_rate)
stop_min <- min(scaled_data$stop_rate)
stop_range <- stop_max - stop_min
scaled_data$stop_rate <- (scaled_data$stop_rate - stop_min) / stop_range

tackle_max <- max(scaled_data$tackle_rate)
tackle_min <- min(scaled_data$tackle_rate)
tackle_range <- tackle_max - tackle_min
scaled_data$tackle_rate <- (scaled_data$tackle_rate - tackle_min) / tackle_range

assist_max <- max(scaled_data$assist_rate)
assist_min <- min(scaled_data$assist_rate)
assist_range <- assist_max - assist_min
scaled_data$assist_rate <- (scaled_data$assist_rate - assist_min) / assist_range

hit_max <- max(scaled_data$hit_rate)
hit_min <- min(scaled_data$hit_rate)
hit_range <- hit_max - hit_min
scaled_data$hit_rate <- (scaled_data$hit_rate - hit_min) / hit_range

hurry_max <- max(scaled_data$hurry_rate)
hurry_min <- min(scaled_data$hurry_rate)
hurry_range <- hurry_max - hurry_min
scaled_data$hurry_rate <- (scaled_data$hurry_rate - hurry_min) / hurry_range

sack_max <- max(scaled_data$sack_rate)
sack_min <- min(scaled_data$sack_rate)
sack_range <- sack_max - sack_min
scaled_data$sack_rate <- (scaled_data$sack_rate - sack_min) / sack_range

# Set final df
final_data <- scaled_data
final_data$pr_comp <- (final_data$hit_rate+final_data$hurry_rate+final_data$sack_rate)/3
final_data$pr_comp <- final_data$pr_comp / max(final_data$pr_comp) * 100 
final_data$rd_comp <- (final_data$stop_rate+final_data$tackle_rate+final_data$assist_rate)/3
final_data$rd_comp <- final_data$rd_comp / max(final_data$rd_comp) * 100 
reduced_cols <- c('player','games','grade_tot','grade_pr','grade_rd','pr_comp','rd_comp')
final_data <- select(final_data, reduced_cols)
final_data <- final_data[order(-final_data$grade_tot),]
final_data[,-1] <- round(final_data[,-1],2)

```


```{r echo=FALSE}
library(knitr)
library(reactable)
BuYlRd <- function(x) rgb(colorRamp(c("#e6fff2", "#80ffbc"))(x), maxColorValue = 255)
reactable(
  final_data,
  pagination = TRUE,
  striped = TRUE,
  highlight = TRUE,
  wrap = FALSE,
  searchable = TRUE,
  style = list(fontFamily = "Work Sans, sans-serif", fontSize = "14px"),
  columns = list(
    player = colDef(name = "Player Name"),
    games = colDef(name = "Games"),
    grade_tot = colDef(name = "PFF Overall"),
    grade_pr = colDef(name = "PFF Pass"),
    grade_rd = colDef(name = "PFF Run"),
    pr_comp = colDef(name = "Composite Pass", style = function(value){
      normalized <- (value - min(final_data$pr_comp)) / (max(final_data$pr_comp) -   min(final_data$pr_comp))
      color <- BuYlRd(normalized)
      list(background = color)
    }),
    rd_comp = colDef(name = "Composite Run", style = function(value){
      normalized <- (value - min(final_data$rd_comp)) / (max(final_data$rd_comp) -   min(final_data$rd_comp))
      color <- BuYlRd(normalized)
      list(background = color)
    })
  ))
```
<br />
<br />
**Players by Pass Defense and Run Defense Scores**  
**PFF and Composite Values**  
The scatterplots below show the relative ranks of each player based on both PFF grades (purple) and composite scores (orange). Pass rush ability moves along the y-axis, while run defense moves along the x-axis. As we move up and to the right within each plot, we find players who are better at each skill.

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)
library(ggplot2)
library(gridExtra)

# Get data for combined grades
final_data$comp_avg <- (final_data$rd_comp + final_data$pr_comp) / 2
final_data$pff_avg <- (final_data$grade_pr + final_data$grade_rd) / 2

p <- ggplot(final_data, aes(x = pr_comp, y= rd_comp, Player=player)) +
  geom_point(color='#ffbc80') +
  xlab("Comp Pass Grade") +
  ylab("Comp Run Grade")

p1 <- ggplotly(p,tooltip = c("Player"))

p2 <- ggplot(final_data, aes(x = grade_pr, y= grade_rd, Player=player)) + 
  geom_point(color='#bc80ff') +
  xlab("PFF Pass Grade") +
  ylab("PFF Run Grade")

p3 <- ggplotly(p2,tooltip = c("Player"))

p4 <- ggplot(final_data, aes(x = pff_avg, y= comp_avg, Player=player)) + 
  geom_point(color='violetred4') +
  xlab("PFF Grade") +
  ylab("Composite Grade")

p5 <- ggplotly(p4,tooltip = c("Player"))

p6 <- subplot(p3, p1, titleX = TRUE, titleY = TRUE, margin = 0.04)
p6 <- p6 %>% layout(autosize = F, width = 800, height = 500)

ppc <- htmltools::div(p6, align="center" )
p6
```
<br />  
**Player Score Comparisons**  
**PFF versus Composite**  
Our two categories – PFF Grades and Composite Scores – capture different items. PFF grades are subjective and more wholistic, while composite scores measure pure production agnostic to game situation. As such, a lot can be gained from analyzing how players score within each category. The chart below plots the two scores for each player, with the purple dot corresponding to the PFF Grade and the orange dot corresponding to the Composite Score. Players with large distances between the two points have a considerable difference in how they are valued by the two metrics.    
  

```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2) 
library(ggalt)   
library(tidyverse)

# Get data for combined grades
db_data <- final_data
db_data$comp_avg <- (db_data$rd_comp + db_data$pr_comp) / 2
db_data$pff_avg <- (db_data$grade_pr + db_data$grade_rd) / 2

db_data <- db_data %>%
   group_by(player) %>%
   summarise(comp_avg = mean(comp_avg), 
             pff_avg = mean(pff_avg))

db_data$comp_avg <- (db_data$comp_avg - min(db_data$comp_avg))/(max(db_data$comp_avg)-min(db_data$comp_avg)) * 100

db_data$pff_avg <- (db_data$pff_avg - min(db_data$pff_avg))/(max(db_data$pff_avg)-min(db_data$pff_avg)) * 100

db_data <- select(db_data, player, comp_avg, pff_avg)
db_data$diff <- (db_data$comp_avg  - db_data$pff_avg)

# Rename columns
names(db_data)[names(db_data) == "player"] <- "Player"
names(db_data)[names(db_data) == "comp_avg"] <- "Comp"
names(db_data)[names(db_data) == "pff_avg"] <- "PFF"

# Make plot
db <- ggplot(db_data, aes(x=Comp, xend=PFF, y=Player)) + 
        #create a thick line between x and xend instead of using defaut 
        #provided by geom_dubbell
        geom_segment(aes(x=Comp, 
                         xend=PFF, 
                         y=Player, 
                         yend=Player), 
                     color="#b2b2b2", size=1.5)+
        geom_dumbbell(color="light blue", 
                      size_x=3.5, 
                      size_xend = 3.5,
                      #Note: there is no US:'color' for UK:'colour' 
                      # in geom_dumbbel unlike standard geoms in ggplot()
                      colour_x="#ffbc80", 
                      colour_xend = "#bc80ff") +
        labs(x=NULL, y=NULL)

db
```
  
  
**Player Radar Charts**  
Now we will look to combine our metrics and see how each player compares to the entire pool. Below are radar charts for each player, where the shaded orange area represents the group average, and the purple lines represent the individuals score in five different categories: Total PFF Grade (TOT), Pass Rush PFF Grade (GP), Run Defense PFF Grade (GR), Pass Rush Composite Grade (CP) and Run Defense Composite Grade (CR). The closer to the edge, the better a player is for a given category.  
<br />
```{r, fig.align="center", echo=FALSE, message=FALSE, warning=FALSE}
library(fmsb) 
library(tibble)

# RADARS
radar_data <- final_data[ , -which(names(final_data) %in% c("games","comp_avg","pff_avg"))]

# Scale PFF ratings
radar_data$grade_tot <- (radar_data$grade_tot - min(radar_data$grade_tot))/(max(radar_data$grade_tot)-min(radar_data$grade_tot)) * 100

radar_data$grade_pr <- (radar_data$grade_pr - min(radar_data$grade_pr))/(max(radar_data$grade_pr)-min(radar_data$grade_pr)) * 100

radar_data$grade_rd <- (radar_data$grade_rd - min(radar_data$grade_rd))/(max(radar_data$grade_rd)-min(radar_data$grade_rd)) * 100

radar_data <- column_to_rownames(radar_data, var = "player")

# Define the variable ranges: maximum and minimum
max_min <- data.frame(
  grade_tot = c(100, 0, mean(radar_data$grade_tot)),
  grade_pr = c(100, 0, mean(radar_data$grade_pr)),
  grade_rd = c(100, 0, mean(radar_data$grade_rd)),
  pr_comp = c(100, 0, mean(radar_data$pr_comp)),
  rd_comp = c(100, 0, mean(radar_data$rd_comp))
)

rownames(max_min) <- c("Max", "Min", "Average")

df <- rbind(max_min, radar_data)

# Helper Function
create_beautiful_radarchart <- function(data, color = "#00AFBB", 
                                        vlabels = colnames(data), vlcex = 0.7,
                                        caxislabels = NULL, title = NULL, ...){
  radarchart(
    data, axistype = 1,
    # Customize the polygon
    pcol = color, pfcol = scales::alpha(color, 0.5), plwd = 2, plty = 1,
    # Customize the grid
    cglcol = "grey", cglty = 1, cglwd = 0.8,
    # Customize the axis
    axislabcol = "grey", 
    # Variable labels
    vlcex = vlcex, vlabels = vlabels,
    caxislabels = caxislabels, title = title, ...
  )
}


# Edit DF
names(df)[names(df) == "grade_tot"] <- "TOT"
names(df)[names(df) == "grade_pr"] <- "GP"
names(df)[names(df) == "grade_rd"] <- "GR"
names(df)[names(df) == "rd_comp"] <- "CR"
names(df)[names(df) == "pr_comp"] <- "CP"

opar <- par() 
# Define settings for plotting in a 3x4 grid, with appropriate margins:
par(mar = rep(1.25,4.5))
par(mfrow = c(4,5))
# Produce a radar-chart for each student
for (i in 4:nrow(df)) {
  radarchart(
    df[c(1:3, i), ],
    pfcol = c(scales::alpha("#ffbc80", 0.5),scales::alpha("#e4cdff", 0.5)),
    pcol= c(NA, "#bc80ff"), plty = 1, plwd = 2, cglty=1, cglcol=c("gray85"),
    title = row.names(df)[i]
  )
}
# Restore the standard par() settings
par <- par(opar) 
```
  
### **Insights**  
  
**Scatter Plots**  
Looking at the scatter plot of PFF grades, we can see that run defense has a lot more concentration near the top than does pass rush. This highlights the notion of skill scarcity and shows that elite pass rushing is much harder to find than elite run stopping.  
  
**Score Comparisons**  
The extreme cases within the player comparison visual are the most interesting as they highlight instances where there is a significant difference in player value. Patrick Jones II, Osa Odighizuwa, and Rashad Weaver are three examples of players who are ranked much more favorable by PFF metrics versus the comparative scores. Comparatively, Gregory Rousseau, Joseph Ossai, and Hamlicar Rashed Jr. are all ranked much more favorably by the comparative scores versus the PFF metrics.  
  
**Player Radar Charts**  
The radar charts highlight the diversity of each player. Players with large purple polygons (such as Chris Rumph II, Quncy Roche, and Jordan Smith) are consistently near the top of each metric, whereas players with small areas are near the bottom. Interesting cases arise where players have one peak, such as Osa Odighizuwa, Cameron Sample, and Hamlicar Rashed Jr. These players all have a specific area that they excel at but are near the middle of the pack in every other category.  
  
### **Future Iterations**  
This analysis is a quick glance at a very specific subset of players. It is by no means exhaustive and will benefit from more and better data. Additionally, future research will allow for more accurately weighted models which will enhance the predictive value of the comparative scores.  
  
***